---
title: "BCT Favorability & Co-Walking Analysis — Methods Demonstration"
author: "Sofia Roth"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, fig.width = 10, fig.height = 5.5)
set.seed(2026)
```

> **Note on data:** Participant data from this NIH-funded study (P30AG063786)
> is confidential. This notebook demonstrates the *exact analysis pipeline* I
> ran during my internship at the Feinstein Institutes for Medical Research,
> applied to **simulated data** that matches the study's structure (N = 146
> caregivers, two trial arms, Likert-scale survey responses). Simulated
> results will differ from the published findings summarized in the poster.

```{r libraries}
library(tidyverse)
library(psych)       # Cronbach's alpha, descriptives
library(rstatix)     # Kendall's W
library(effectsize)  # rank-biserial effect sizes
library(boot)        # BCa bootstrapped CIs
library(patchwork)   # multi-panel figures
```

## 1. Simulating the study structure

The real dataset contained a post-study survey gathering responses from 146 caregivers
of persons with Alzheimer's disease and AD-related dementias (AD/ADRD) randomized to a 1-hour daily walking trial (n = 69) or a 4-hour hourly walking trial (n = 77).
Each participant rated four Behavior Change Techniques (BCTs) such as Goal Setting, Action Planning, Prompt/Cue, Self-Monitoring on three items each
(satisfaction, ease of use, helpfulness), on a 1–4 Likert scale.

I simulate that structure here, building in a slight preference for Prompt/Cue to mimic the direction of the real finding.

```{r simulate-rq1}
n <- 146

simulate_bct <- function(n, base_probs) {
  replicate(3, sample(1:4, n, replace = TRUE, prob = base_probs))
}

rq1_df <- tibble(
  record_id   = 1:n,
  trial_group = c(rep("1", 69), rep("4", 77))
)

# Prompt/Cue gets a distribution shifted slightly toward higher ratings
bct_probs <- list(
  goal   = c(.05, .15, .45, .35),
  action = c(.05, .18, .45, .32),
  cue    = c(.03, .10, .42, .45),
  self   = c(.06, .16, .44, .34)
)

for (bct in names(bct_probs)) {
  items <- simulate_bct(n, bct_probs[[bct]])
  rq1_df[[paste0(bct, "_sat")]]  <- items[, 1]
  rq1_df[[paste0(bct, "_ease")]] <- items[, 2]
  rq1_df[[paste0(bct, "_help")]] <- items[, 3]
}
```

## 2. Scale reliability: Cronbach's alpha

Before combining the three items per BCT into a composite score, I checked
internal consistency. If the items don't relate to another, a composite isn't
justified.

```{r alpha}
alpha_goal <- psych::alpha(rq1_df %>% select(goal_sat, goal_ease, goal_help))
alpha_cue  <- psych::alpha(rq1_df %>% select(cue_sat, cue_ease, cue_help))
alpha_goal$total$raw_alpha
alpha_cue$total$raw_alpha
```

*(In the real data, alphas supported compositing passing a significance level of greater than 0.70. Simulated items here are
independent draws, so alpha may be low)*

## 3. Composite scores and descriptives

```{r composites}
mean_df <- rq1_df %>%
  mutate(
    goal_overall   = rowMeans(pick(goal_sat, goal_ease, goal_help), na.rm = TRUE),
    action_overall = rowMeans(pick(action_sat, action_ease, action_help), na.rm = TRUE),
    cue_overall    = rowMeans(pick(cue_sat, cue_ease, cue_help), na.rm = TRUE),
    self_overall   = rowMeans(pick(self_sat, self_ease, self_help), na.rm = TRUE)
  ) %>%
  select(record_id, trial_group, ends_with("_overall"))

psych::describe(mean_df %>% select(ends_with("_overall")))
```

## 4. RQ1a: Which BCT do caregivers favor?

**Analyst decision:** the composites are built from ordinal Likert items and
are not normally distributed, and each participant rated all four BCTs
(repeated measures). That rules out repeated-measures ANOVA in favor of the
**Friedman test**, with **Kendall's W** as the effect size.

```{r friedman}
friedman_data <- mean_df %>%
  select(ends_with("_overall")) %>%
  drop_na()

friedman.test(as.matrix(friedman_data))

long_df <- mean_df %>%
  drop_na() %>%
  pivot_longer(ends_with("_overall"), names_to = "BCT", values_to = "Score")

friedman_effsize(long_df, Score ~ BCT | record_id)
```

### Post-hoc pairwise comparisons

With six pairwise comparisons, the family-wise error rate multiplies — so each
Wilcoxon signed-rank p-value is **Bonferroni-corrected** (× 6, capped at 1).
I also report **rank-biserial r** with 95% CIs, because a p-value wont report the magnitude of the difference.

```{r pairwise}
bct_labels <- c(goal_overall = "Goal Setting", action_overall = "Action Planning",
                cue_overall = "Prompt/Cue", self_overall = "Self-Monitoring")

pairs <- combn(names(bct_labels), 2, simplify = FALSE)

pairwise_results <- map_dfr(pairs, function(p) {
  d <- mean_df %>% select(all_of(p)) %>% drop_na()
  test <- wilcox.test(d[[p[1]]], d[[p[2]]], paired = TRUE, exact = FALSE)
  es   <- rank_biserial(d[[p[1]]], d[[p[2]]], paired = TRUE, ci = 0.95)
  tibble(
    comparison = paste(bct_labels[[p[1]]], "vs", bct_labels[[p[2]]]),
    V = test$statistic, p_raw = test$p.value,
    p_adj = pmin(test$p.value * 6, 1),
    r_rb = round(es$r_rank_biserial, 3),
    CI = sprintf("[%.2f, %.2f]", es$CI_low, es$CI_high)
  )
})

pairwise_results
```

## 5. RQ1b: Does favorability differ by the 1-hour trial vs 4-hour trial?

Between-group comparison of independent samples with ordinal outcomes therefore
**unpaired Wilcoxon (Mann-Whitney U)**, Bonferroni-corrected across the four
BCTs.

```{r trial-comparison}
trial_results <- map_dfr(names(bct_labels), function(bct) {
  d <- mean_df %>% select(trial_group, score = all_of(bct)) %>% drop_na()
  test <- wilcox.test(score ~ trial_group, data = d, exact = FALSE)
  es   <- rank_biserial(score ~ trial_group, data = d, ci = 0.95)
  tibble(
    BCT = bct_labels[[bct]],
    W = test$statistic, p_raw = test$p.value,
    p_adj = pmin(test$p.value * 4, 1),
    r_rb = round(es$r_rank_biserial, 3)
  )
})

trial_results
```

```{r violin-plot, fig.height=7, fig.width=12}
plot_df <- long_df %>%
  mutate(
    BCT = factor(bct_labels[BCT], levels = bct_labels),
    trial_group = factor(trial_group, levels = c("1", "4"),
                         labels = c("Trial 1 (1-hour)", "Trial 2 (4-hour)"))
  )

ggplot(plot_df, aes(BCT, Score, fill = trial_group)) +
  geom_violin(position = position_dodge(.8),
              alpha = .25, color = NA) +
  geom_boxplot(position = position_dodge(.8),
               width = .25, outlier.shape = NA, alpha = .9) +
  stat_summary(fun = mean, geom = "point",
               shape = 23, size = 4,
               fill = "white", color = "black",
               position = position_dodge(.8)) +
  scale_fill_manual(values = c("#0072B2", "#D55E00")) +
  scale_y_continuous(limits = c(1, 4), breaks = 1:4) +
  labs(
    title = "Ratings Were Similar Across Trials",
    x = "Behavior change techniques",
    y = "BCT composite rating",
    fill = "Trial"
  ) +
  theme_minimal(base_size = 16) +
  theme(
    plot.title = element_text(face = "bold", size = 22),
    axis.title = element_text(size = 14),
    legend.position = "top",
    panel.grid.major.x = element_blank()
  )
```

## 6. RQ2: Co-walking frequency and perceived physical and emotional health

The co-walking survey was an exploratory, late-stage measure (n = 86 of 146 in
the real study). Added to the post-study survey to explore spillover effects of caregiver physical activity.  **Analyst decisions:** (1) the research question concerns
whether the *frequency* of co-walking relates to perceived improvements, so
"Never" responses (caregivers who did not co-walk at all) were excluded to
isolate the frequency gradient among participants who co-walked; (2) both
variables are ordinal therefore **Spearman's rho**; (3) heavy ties in 5-point scales so implemented 
`exact = FALSE`; (4) modest exploratory subsample leads to not trust asymptotic CIs; compute **bias-corrected and accelerated (BCa) bootstrap CIs** (5,000 resamples) instead.

```{r simulate-rq2}
n2 <- 86
latent <- rnorm(n2)
walking_df <- tibble(
  activity_together  = pmin(pmax(round(2.8 + 1.1 * latent + rnorm(n2, 0, .8)), 1), 5),
  positive_physical  = pmin(pmax(round(3.0 + 0.9 * latent + rnorm(n2, 0, .9)), 1), 5),
  positive_emotional = pmin(pmax(round(3.1 + 0.8 * latent + rnorm(n2, 0, .9)), 1), 5)
)

# Exclude "Never" responses (activity_together == 1)
walking_df <- walking_df %>% filter(activity_together != 1)
cat("N after excluding 'Never':", nrow(walking_df), "\n")
```

```{r spearman}
cor.test(walking_df$activity_together, walking_df$positive_physical,
         method = "spearman", exact = FALSE)

spearman_stat <- function(data, indices, yvar) {
  d <- data[indices, ]
  cor(d$activity_together, d[[yvar]], method = "spearman")
}

set.seed(123)
boot_physical <- boot(walking_df, spearman_stat, R = 5000, yvar = "positive_physical")
boot.ci(boot_physical, type = "bca")

set.seed(123)
boot_emotional <- boot(walking_df, spearman_stat, R = 5000, yvar = "positive_emotional")
boot.ci(boot_emotional, type = "bca")
```

```{r bubble-plot, fig.width=16, fig.height=6}
make_bubble <- function(data, yvar, col, panel) {

  yvar_sym <- enquo(yvar)

  d <- data %>%
    filter(!is.na(activity_together), !is.na(!!yvar_sym))

  test_result <- cor.test(
    d$activity_together,
    pull(d, !!yvar_sym),
    method = "spearman",
    exact = FALSE
  )

  rho  <- test_result$estimate
  pval <- test_result$p.value

  rho_label <- paste0("\u03C1 = ", sprintf("%.2f", rho))
  p_label   <- if (pval < 0.001) "p < .001" else paste0("p = ", sprintf("%.3f", pval))

  d %>%
    count(activity_together, !!yvar_sym) %>%
    ggplot(aes(activity_together, !!yvar_sym, size = n)) +
    geom_smooth(method = "lm", se = FALSE, linetype = "dashed",
                color = col, linewidth = 1.1) +
    geom_point(color = col, alpha = 0.55) +
    scale_size_area(max_size = 22, guide = "none") +
    scale_x_continuous(
      breaks = 1:5,
      labels = c("Never", "Rarely", "Sometimes", "Often", "Almost\nAlways")
    ) +
    scale_y_continuous(breaks = 1:5, limits = c(0.5, 5.5)) +
    annotate("text", x = 4.8, y = 1.15, label = rho_label,
             hjust = 1, size = 5, fontface = "italic", color = "grey30") +
    annotate("text", x = 4.8, y = 0.75, label = p_label,
             hjust = 1, size = 5, fontface = "italic", color = "grey30") +
    labs(
      x = "Co-walking frequency",
      y = "Perceived improvement (1\u20135)",
      title = panel
    ) +
    theme_minimal(base_size = 18) +
    theme(
      plot.title = element_text(face = "bold", color = col),
      axis.title.x = element_text(size = 16, margin = margin(t = 10)),
      axis.title.y = element_text(size = 16, margin = margin(r = 10))
    )
}

pa <- make_bubble(walking_df, positive_physical, "#003CA5", "a) Physical health")
pb <- make_bubble(walking_df, positive_emotional, "#00823D", "b) Emotional health")

(pa | pb) +
  plot_annotation(
    title = "More Frequent Walking Together Was Associated With Better Perceived Health",
    theme = theme(plot.title = element_text(face = "bold", size = 22))
  ) &
  theme(plot.margin = margin(5.5, 10, 5.5, 10))
```

## Real study results conclusion 

On the actual data, this pipeline showed: **Prompt/Cue BCT rating suggest direct prompts are more favored by AD/ADRD caregivers.** (rated significantly higher than Action Planning, p < .05). Although, BCT favorability has **no
significant difference** between the 1-hour and 4-hour trial. This did not support HQ1b representing message burden not driving ratings. The **strong association of co-walking frequency** with care recipient physical (rho = 0.66) and emotional (rho = 0.63) health, both p < .001 with BCa CIs excluding zero.

These insights will drive future research and intervention
design development for caregivers and patients with
AD/ADRD. The consistency of BCT favorability across both
trials demonstrates that text-based interventions remain
effective in both one-hour and four-hour formats, offering
flexibility to accommodate caregivers’ unique demands and
schedules. Future research should leverage satisfaction data
to optimize BCT methods and co-walking data to further
investigate the therapeutic benefits of shared physical activity
on patient well-being.

This work was supported by the National Institute on Aging of the
National Institutes of Health under Award Number P30AG063786.
We thank all participating caregivers for their valuable
contributions.
